Adhesion, hydrodynamics and rheology of vesicle doublets

### 2019-02-10 vesicle_code/docs/adhesion/adhesion.tex
edits to intro

\title{Hydrodynamics and rheology of an adhesive vesicle doublet}
% Force line breaks with \\
%\thanks{A footnote to the article title}%

@@

\begin{abstract}
The dynamics of an adhesive vesicle doublet under various flowing conditions is
investigated numerically using a high-order adaptive-in-time boundary
integral equation formulation. In a quiescent flow, two nearby vesicles
move slowly towards each other under the adhesive potential, pushing out
fluid between them to form a vesicle doublet at equilibrium. A lubrication
analysis on such draining of a thin film gives the dependencies of
draining time on adhesion strength and separation distance that are in
good agreement with numerical results.
%
In a microfluid trap where the stagnation of an extensional flow is dynamically
placed in the middle of a vesicle doublet through an active control
loop, novel dynamics of a vesicle doublet is observed. Numerical
simulations show that there exists a critical extensional flow rate
above which adhesive interaction is overcome by the 
diverging stream, thus providing a simple method to measure the adhesion
strength between two vesicle membranes. 
%
In a planar shear flow, numerical simulations reveal that a vesicle
doublet may form as two vesicles approach each other provided that the adhesion
strength is sufficiently large at a given vesicle reduced area. Once a
doublet is formed, the effective shear viscosity of a dilute suspension
of vesicle doublets is found to be a function of both reduced area and
the adhesion strength.  Results from these numerical studies and
analysis shed light on the hydrodynamic effects of adhesive interactions
between vesicles in a viscous fluid.
%\begin{description}
%\item[Usage]
%Secondary publications and information retrieval purposes.
%\item[PACS numbers]
%May be entered using the \verb+\pacs{#1}+ command.
%\item[Structure]
%You may use the \texttt{description} environment to structure your abstract;
%use the optional argument of the \verb+\item+ command to give the category of each item. 
%\end{description}
\end{abstract}

@@

\section{Introduction}
Vesicles (closed fluid-filled phospholipid bilayer membranes) have been
widely utilized as biological cell mimics in biophysics and material
engineering~\cite{sackmann1996, Barthes-Biesel2016_ARFM}.  The ever
expanding applications have encouraged detailed experimental,
theoretical, and numerical investigations of vesicle hydrodynamics in
applied flow, electric fields, and magnetic fields.  Theoretical
investigations often take the form of small-deformation analysis or
spheroidal dynamics of a nearly-spherical vesicle (or nearly-circular in
two dimensions) subjected to an applied flow or under an external
forcing~\cite{Barthes-BieselRallison1981_JFM, Misbah2006_PRL,
Vlahovska2007_PRE, Finken2008_EPL, ZhangZahnTanLin2013_PoF,
Nganguia2013_PRE}.  On the other hand, various numerical methods have
been developed for simulating the transient hydrodynamics of vesicle
suspensions~\cite{BagchiJohoson2005_JBE, Biben2005_EJP,
Veerapaneni2009_JCP, SeolHuKimLai2016_JCP}.

Hydrodynamics of a single vesicle in Stokes flow has been extensively
investigated. In a planar shear flow, the vesicle hydrodynamics is
characterized by the reduced volume (reduced area in two dimensions),
viscosity contrast between interior and exterior fluids, and shear rate
of the imposed far-field fluid flow. In addition, a vesicle with a rigid
particle inside is also investigated as a biological mimic of a
eukaryotic cell with a nucleus that occupies nearly half of the
intracellular volume~\cite{Veerapaneni2011_PRL}. Small-deformation
analysis shows that a vesicle tank-treads in a planar shear flow for low
viscosity contrast and shear rate. At high viscosity contrast the
tank-treading dynamics transitions to tumbling
dynamics~\cite{Misbah2006_PRL, Vlahovska2007_PRE}, and this leads to a
transition in effective shear viscosity of the vesicle
suspension~\cite{Misbah2006_PRL,Vitkova2008_BJ} that is also validated
by direct numerical simulations~\cite{GhigliottiBibenMisbah2010_JFM} and
experiments~\cite{DeschampsKantsler2009_PNAS,
KantslerSegreSteinberg2008_EPL, ZabuskySegreDeschamps2011_PoF}.  Between
tank-treading and tumbling vesicle dynamics, a breathing (tremble) mode
is also observed~\cite{Misbah2006_PRL, KantslerSegreSteinberg2008_PRL,
ZhaoShaqfeh2011_JFM, SpannZhaoShaqfeh2014_PoF} to affect the effective
shear viscosity in a different way.  In an extensional flow (planar or
uniaxial), vesicle shape dynamics depends sensitively on the vesicle
reduced volume and the elastic capillary
number~\cite{KantslerSegreSteinberg2008_PRL, ZhaoShaqfeh2013_JFM,
Narsimhan2014_JFM, DahlNarsimhanGouveia2016_SoftMatt}: Asymmetric shape
and oscillatory undulation of the vesicle membrane are two examples of
the complex vesicle hydrodynamics in an extensional flow.

Collective hydrodynamics of vesicles is dictated by the vesicle-vesicle
interactions.  In a quiescent flow, vesicle-vesicle adhesion leads to
the formation of vesicle doublets~\cite{Ziherl2007_PRL,
ZiherlSvetina2007_PNAS} or
clusters~\cite{SvetinaZiherl2008_Bioelectrochemistry,
FlormannAouane2017_SciReports}.  As a model for red blood cell (RBC)
aggregates, an idealized model for adhesive vesicle-vesicle interactions
gives rise to vesicle configurations similar to those observed in
experiments of fibrinogen-induced RBC
aggregates~\cite{SvetinaZiherl2008_Bioelectrochemistry,
FlormannAouane2017_SciReports}.  Using the Lennard-Jones
(L.-J.)~potential for point-point interaction between two vesicle
membranes without the molecular details of adhesive interactions between
RBCs, Flormann {\em et al.}~found the contact zone to buckle under
strong adhesion~\cite{FlormannAouane2017_SciReports}.  A direct
consequence of the buckling is the sigmoidal shape of the adhesion
region, which is also observed in RBC doublets and explained in a
slightly different model~\cite{ZiherlSvetina2007_PNAS}.  It remains
unknown how adhesive interactions between vesicles might lead to
different vesicle hydrodynamics that results in different rheological
properties of a vesicle suspension. Such studies will provide useful
insight to the rheology of blood flow and how to use nano solutes to
control the hypertension by adjusting the adhesive interactions between
RBCs.

In numerical simulations of a vesicle suspension, the adhesive
interaction between two vesicles is often
ignored~\cite{Veerapaneni2009_JCP,  RahimianVeerapaneniBiros2010_JCP}
for numerical convenience, mostly because a small time step is often
needed to resolve the dynamics of the lubrication thin film between two
membranes with adhesion.  However, adhesive interactions between RBCs
lead to different RBC clusters that are expected to alter the
rheological properties of the RBC suspension~\cite{NeuMeiselman2002_BJ,
SvetinaZiherl2008_Bioelectrochemistry}. Hydrodynamic simulations of
adhesive vesicles in a quiescent flow have revealed physical insights to
the equilibrium shapes of RBC aggregates in both
experiments~\cite{FlormannAouane2017_SciReports} and
theory~\cite{ZiherlSvetina2007_PNAS}. The main goal of this work is to
investigate the hydrodynamics of adhesive vesicles in flowing conditions
common in microfluidic vesicle flow, such as a planar shear flow and an
extensional flow.

The electrostatic nature of lipid molecules (a hydrophobic tail and a
hydrophilic head with an electric dipole) complicates the interactions
between a lipid bilayer membrane and another bilayer
membrane~\cite{EvansMetcalfe1984_BJ, Book_PhysicalBasisCellAdhesion,
Book_IntermolecularSurfaceForces, PerutkovaFrank-Bertoncelij2013_CSB} or
a solid (such as a glass substrate or a nano particle).  Adhesion
between a vesicle and a substrate has been extensively studied with more
focuses on the static equilibrium shapes~\cite{Seifert1990_PRA,
ShiFengGao2006_ActaMechSin, LinFreund2007_IntJSolidsStructures,
GruhnFrankeDimova2007_Langmuir,das2008adhesion, zhang2009phase,
SteinkuhlerAgudo-Canalejo2016_BJ} than the transient adhesion
process~\cite{cantat1999lift, suk-sei2001, BlountMiksisDavis2013_PRSa}.
Adhesive interactions between lipid membranes are essential in many
biomedical, biological, and biophysical processes.  For example, vesicle
adhesion is crucial to initiate membrane fusion and fission in
endocytosis and exocytosis, the transport of small vesicles through
membrane surfaces.  In the absence of an external electric field and
ions in the solvents, it is reported that the adhesive interactions
between two membranes can be well approximated by the L.-J.-type
potential~\cite{FlormannAouane2017_SciReports}, that consists of a
long-range attraction component and a short-range repulsion
component~\cite{Book_IntermolecularSurfaceForces}, and the combination
of the two gives rise to adhesion at the separation distance that
minimizes the interaction
potential~\cite{Book_IntermolecularSurfaceForces}.

@@

Boundary integral equation (BIE) approaches are well-suited for solving
the low Reynolds flow problems considered here as they lead to reduction
in dimensionality and achieve high-order accuracy even for moving
geometry problems. The BIE resolves the self-vesicle interaction using
quadrature for logarithmically singular integrands~\cite{alp1999}, but
as the vesicles adhere, the BIE solver must also resolve the
vesicle-vesicle {\em nearly-singular} hydrodynamic forces.  We use an
interpolation-based quadrature rule~\cite{qua-bir2014} to maintain
high-order accuracy for all vesicle-vesicle separation distances.  To
overcome the numerical stiffness induced by the membrane bending forces
and to control the global error, we employ a high-order spectrally
deferred implicit-explicit adaptive time stepping
scheme~\cite{quaife2016adaptive}.

@@

\begin{figure}[htp]
\centering
\includegraphics[height=0.4\textwidth]{figs/relaxationManyVesTension.png}
\includegraphics[height=0.4\textwidth]{figs/relaxationManyVesAdhesion.png}
\caption{\label{fig:relaxationManyVesAdhesionTension} The tension ({\em
left}) and adhesion force ({\em right}) of the final configuration from
Figure~\ref{fig:relaxationManyVes}.  Along the bounding vesicle, the
tension is largest and the adhesion is smallest in the regions that are
well-separated from the interior vesicles.  Similar to other
simulations, the interior vesicles are relatively flat and have a large
negative tension in the contact regions.}
\end{figure}

>>>>

\title{Adhesion, hydrodynamics and rheology of vesicle doublets}
% Force line breaks with \\
%\thanks{A footnote to the article title}%

@@

\begin{abstract}
The dynamics of an adhesive vesicle doublet under various flow conditions is
investigated numerically using a high-order, adaptive-in-time boundary
integral method. In a quiescent flow, two nearby vesicles
move slowly towards each other under the adhesive potential, pushing out
fluid between them to form a vesicle doublet at equilibrium. A lubrication
analysis on such draining of a thin film gives the dependencies of
draining time on adhesion strength and separation distance that are in
good agreement with numerical results.
%
In a microfluid trap where the stagnation of an extensional flow is dynamically
placed in the middle of a vesicle doublet through an active control
loop, novel dynamics of a vesicle doublet is observed. Numerical
simulations show that there exists a critical extensional flow rate
above which adhesive interaction is overcome by the 
diverging stream, thus providing a simple method to measure the adhesion
strength between two vesicle membranes. 
%
In a planar shear flow, numerical simulations reveal that a vesicle
doublet may form as two vesicles approach each other provided that the adhesion
strength is sufficiently large at a given vesicle reduced area. Once a
doublet is formed, the effective shear viscosity of a dilute suspension
of vesicle doublets is found to be a function of both reduced area and
the adhesion strength.  Results from these numerical studies and
analysis shed light on the hydrodynamic effects of adhesive interactions
between vesicles in a viscous fluid.
%\begin{description}
%\item[Usage]
%Secondary publications and information retrieval purposes.
%\item[PACS numbers]
%May be entered using the \verb+\pacs{#1}+ command.
%\item[Structure]
%You may use the \texttt{description} environment to structure your abstract;
%use the optional argument of the \verb+\item+ command to give the category of each item. 
%\end{description}
\end{abstract}

@@

\section{Introduction}
Vesicles (closed fluid-filled phospholipid bilayer membranes) have been
widely utilized as biological cell mimics in biophysics and material
engineering~\cite{sackmann1996, Barthes-Biesel2016_ARFM}.  The ever
expanding applications have encouraged detailed experimental,
theoretical and numerical investigations of vesicle hydrodynamics in
applied flow, electric and magnetic fields.  Theoretical
investigations often take the form of small-deformation analysis or
spheroidal dynamics of a nearly-spherical vesicle (or nearly-circular in
two dimensions) subjected to an applied flow or under an external
forcing~\cite{Barthes-BieselRallison1981_JFM, Misbah2006_PRL,
Vlahovska2007_PRE, Finken2008_EPL, ZhangZahnTanLin2013_PoF,
Nganguia2013_PRE}.  On the other hand, various numerical methods have
been developed for simulating the transient hydrodynamics of vesicle
suspensions e.g., \cite{BagchiJohoson2005_JBE, Biben2005_EJP,
Veerapaneni2009_JCP, SeolHuKimLai2016_JCP}.

Hydrodynamics of a single vesicle in Stokes flow has been extensively
investigated. In a planar shear flow, the vesicle hydrodynamics is
characterized by the reduced volume (reduced area in two dimensions),
viscosity contrast between interior and exterior fluids, and shear rate
of the imposed far-field fluid flow. In addition, a vesicle with a rigid
particle inside is also investigated as a biological mimic of a
eukaryotic cell with a nucleus that occupies nearly half of the
intracellular volume~\cite{Veerapaneni2011_PRL}. Small-deformation
analysis shows that a vesicle tank-treads in a planar shear flow for low
viscosity contrast and shear rate. At high viscosity contrast the
tank-treading dynamics transitions to tumbling
dynamics~\cite{Misbah2006_PRL, Vlahovska2007_PRE}, and this leads to a
transition in effective shear viscosity of the vesicle
suspension~\cite{Misbah2006_PRL,Vitkova2008_BJ} that is also validated
by direct numerical simulations~\cite{GhigliottiBibenMisbah2010_JFM} and
experiments~\cite{DeschampsKantsler2009_PNAS,
KantslerSegreSteinberg2008_EPL, ZabuskySegreDeschamps2011_PoF}.  Between
tank-treading and tumbling vesicle dynamics, a breathing (tremble) mode
is also observed~\cite{Misbah2006_PRL, KantslerSegreSteinberg2008_PRL,
ZhaoShaqfeh2011_JFM, SpannZhaoShaqfeh2014_PoF} to alter the effective
shear viscosity.  In an extensional flow (planar or
uniaxial), vesicle shape dynamics depends sensitively on the vesicle
reduced volume and the elastic capillary
number~\cite{KantslerSegreSteinberg2008_PRL, ZhaoShaqfeh2013_JFM,
Narsimhan2014_JFM, DahlNarsimhanGouveia2016_SoftMatt}: Asymmetric shape
and oscillatory undulation of the vesicle membrane are two examples of
the complex vesicle hydrodynamics in an extensional flow.

Collective hydrodynamics of vesicles is dictated by the vesicle-vesicle
interactions.  In a quiescent flow, vesicle-vesicle adhesion leads to
the formation of vesicle doublets~\cite{Ziherl2007_PRL,
ZiherlSvetina2007_PNAS} or
clusters~\cite{SvetinaZiherl2008_Bioelectrochemistry,
FlormannAouane2017_SciReports}.  As a model for red blood cell (RBC)
aggregates, an idealized model for adhesive vesicle-vesicle interactions
gives rise to vesicle configurations similar to those observed in
experiments of fibrinogen-induced RBC
aggregates~\cite{SvetinaZiherl2008_Bioelectrochemistry,
FlormannAouane2017_SciReports}.  Using the Lennard-Jones
(L.-J.)~potential for point-point interaction between two vesicle
membranes without the molecular details of adhesive interactions between
RBCs, Flormann {\em et al.}~\cite{FlormannAouane2017_SciReports} found the contact zone to buckle under
strong adhesion.  A direct
consequence of the buckling is the sigmoidal shape of the adhesion
region, which is also observed in RBC doublets and explained in a
slightly different model~\cite{ZiherlSvetina2007_PNAS}.  It remains
unknown how adhesive interactions between vesicles might lead to
different vesicle hydrodynamics that results in different rheological
properties of a vesicle suspension. Such studies will provide useful
insight into the rheology of blood flow and how to use nano solutes to
control the hypertension by adjusting the adhesive interactions between
RBCs.

In numerical simulations of a vesicle suspension, the adhesive
interaction between two vesicles is often
ignored~\cite{Veerapaneni2009_JCP,  RahimianVeerapaneniBiros2010_JCP}
for numerical convenience, mostly because a small time step is often
needed to resolve the dynamics of the lubrication thin film between two
membranes with adhesion.  However, adhesive interactions between RBCs
lead to different RBC clusters that are expected to alter the
rheological properties of the RBC suspension~\cite{NeuMeiselman2002_BJ,
SvetinaZiherl2008_Bioelectrochemistry}. Hydrodynamic simulations of
adhesive vesicles in a quiescent flow have revealed physical insights to
the equilibrium shapes of RBC aggregates in both
experiments~\cite{FlormannAouane2017_SciReports} and
theory~\cite{ZiherlSvetina2007_PNAS}. The main goal of this work is to
investigate the hydrodynamics of adhesive vesicles in flow conditions
common in microfluidics such as a planar shear flow and an
extensional flow.

The electrostatic nature of lipid molecules (a hydrophobic tail and a
hydrophilic head with an electric dipole) complicates the interactions
between a lipid bilayer membrane and another bilayer
membrane~\cite{EvansMetcalfe1984_BJ, Book_PhysicalBasisCellAdhesion,
Book_IntermolecularSurfaceForces, PerutkovaFrank-Bertoncelij2013_CSB} or
a solid (such as a glass substrate or a nano particle).  Adhesion
between a vesicle and a substrate has been extensively studied with more
focus on the static equilibrium shapes~\cite{Seifert1990_PRA,
ShiFengGao2006_ActaMechSin, LinFreund2007_IntJSolidsStructures,
GruhnFrankeDimova2007_Langmuir,das2008adhesion, zhang2009phase,
SteinkuhlerAgudo-Canalejo2016_BJ} than the transient adhesion
process~\cite{cantat1999lift, suk-sei2001, BlountMiksisDavis2013_PRSa}.
Adhesive interactions between lipid membranes are essential in many
biomedical, biological, and biophysical processes.  For example, vesicle
adhesion is crucial to initiate membrane fusion and fission in
endocytosis and exocytosis and the transport of small vesicles through
membrane surfaces.  In the absence of an external electric field and
ions in the solvents, it is reported that the adhesive interactions
between two membranes can be well approximated by the L.-J.-type
potential~\cite{FlormannAouane2017_SciReports}, that consists of a
long-range attraction component and a short-range repulsion
component~\cite{Book_IntermolecularSurfaceForces}, and the combination
of the two gives rise to adhesion at the separation distance that
minimizes the interaction
potential~\cite{Book_IntermolecularSurfaceForces}.

@@

Boundary integral equation (BIE) approaches are well-suited for solving
the low Reynolds flow problems considered here as they lead to reduction
in dimensionality and achieve high-order accuracy even for moving
geometry problems. When the vesicles adhere, one major issue for BIE solvers is to resolve the vesicle-vesicle hydrodynamic forces which become {\em nearly singular}. 
We use an interpolation-based quadrature rule~\cite{qua-bir2014} to maintain
high-order accuracy for all vesicle-vesicle separation distances.  To
overcome the numerical stiffness induced by the membrane bending forces
and to control the global error, we employ a high-order spectral
deferred implicit-explicit adaptive time stepping
scheme~\cite{quaife2016adaptive}.

@@

\begin{figure}[htp]
\centering
\fbox{
\includegraphics[height=0.2\textwidth,trim={10cm 0cm 10cm 0cm},clip]    {figs/relaxationManyVesInitialCondition.png}
}
\fbox{
\includegraphics[height=0.2\textwidth]{figs/relaxationManyVesTension.png}
}
\fbox{
\includegraphics[height=0.2\textwidth]{figs/relaxationManyVesAdhesion.png}
}
\caption{\label{fig:relaxationManyVesAdhesionTension} The tension ({\em
left}) and adhesion force ({\em right}) of the final configuration from
Figure~\ref{fig:relaxationManyVes}.  Along the bounding vesicle, the
tension is largest and the adhesion is smallest in the regions that are
well-separated from the interior vesicles.  Similar to other
simulations, the interior vesicles are relatively flat and have a large
negative tension in the contact regions.}
\end{figure}
